Supplement 1: Different Response Distributions

### 2018-09-28 pkg/disttree/inst/draft/supplement/supplement_distributions.Rnw
some touch-ups
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2297 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: cf6ac1fc0569166b41731ec77e0c53af34018b88

\title{Supplement 1: Testing different distributions}

@@

\Abstract{In the application discussed in the paper ''Distributional 
Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain''
the underlying distribution is assumed to be a left-censored at zero Gaussian distribution.
In this supplement the exact same evaluations as in the paper are performed employing
other distribution assumptions. In particular, a left-censored at zero logistic distribution 
and a hurdle-model are applied and their results are compared to those of the left-censored 
at zero Gaussian distribution presented in the paper.}

@@

\section{Introduction}
This application study is supplementary material to the application presented
in Section~3 of the paper.
There, the presented distributional forest model is employed to obtain 
probabilistic precipitation forecasts in a mountainous region assuming that the 
response variable (the power-transformed daily precipitation sum) follows a 
left-censored Gaussian distribution with censoring point~$0$.

To assess how appropriate this distributional assumption is we test the robustness
of the results with respect to two questions: (1)~Are the tails of the distribution
captured appropriately by a Gaussian distribution or are \emph{heavier tails} required?
(2)~Is the point mass at zero (= no rain) driven by the same determinants as the
positive observations (= rain) or should this be captured by a \emph{separate parameter}?
For (1) we compare the censored Gaussian with a \emph{censored logistic} distribution and
for (2) a Gaussian \emph{hurdle model} (or two-part model) is employed instead of the
censored Gaussian.

To compare the performance of the distributional forest to other commonly used
methods the exact same models as in the paper are evaluated on the same data set 
and applying the same settings while only the distribution is exchanged. Further 
details on the models can be found in Section~3.2. of the paper. In particular, 
Table~2 gives an overview over the models and their specifications. As a brief 
reminder the approaches are summarized below.

\begin{itemize}

@@

\item \emph{EMOS} \citep{Gneiting+Raftery+Westveld:2005}: 
Standard ensemble model output statistics models use the ensemble 
mean of total precipitation as regressor in the location submodel 
and the corresponding ensemble standard deviation in the scale submodel.
The parameters are estimated by maximum likelihood.

\item \emph{Prespecified GAMLSS:} Smooth additive splines are selected for
the most relevant predictors based on meteorological expert knowledge
following \cite{Stauffer+Umlauf+Messner:2017}. The model is estimated by maximum
penalized likelihood using a backfitting algorithm (\cite{Stasinopoulos+Rigby:2007}).

\item \emph{Boosted GAMLSS:} Smooth additive splines are selected
automatically from all available variables, using non-cyclic boosting
for parameter estimation (\cite{Hofner+Mayr+Schmid:2016, Messner+Mayr+Zeileis:2017}).
The optimal stopping point for the iteration is evaluated based on 
a (computationally intensive) out-of-bag bootstrap estimate of the log-likelihood.

@@

\section{Distributions applied / Distributional approaches}

The distribution applied in the paper is based on a Gaussian distribution 
for which the probability density function $f$ and the distribution 
function $F$ are defined by
%
\begin{align}
f_{\mathrm{Gaussian}}(Y; \mu, \sigma) &= 
\frac{1}{\sigma} \cdot \phi\left(\frac{Y-\mu}{\sigma}\right) \\
F_{\mathrm{Gaussian}}(Y; \mu, \sigma) &=
\Phi\left(\frac{Y-\mu}{\sigma}\right)
\end{align}
where $\phi$ and $\Phi$ are the probability density function and the 
distribution function of the standard normal distribution $\mathcal{N}(0,1)$.

Alternatively one could also choose a distribution with heavier tails 
such as a logistic distribution with corresponding probability density 
function and distribution function
%
\begin{align}
f_{\mathrm{logistic}}(Y; \mu, \sigma) &= 
\frac{\exp{\left(-\frac{Y - \mu}{\sigma}\right)}}
{\sigma \cdot \left(1 + \exp{\left(-\frac{Y - \mu}{\sigma}\right)}\right)^2} \\
F_{\mathrm{logistic}}(Y; \mu, \sigma) &=
\frac{1}{1+ \exp{\left(-\frac{Y - \mu}{\sigma}\right)}}.
\end{align}
%
Both distributions have to be adapted in our application to capture that
precipitation is non-negative with a point mass at zero (= no rain).
Two common strategies for this are to employ either \emph{left-censoring at zero}
or a \emph{hurdle} (or two-part) approach, combining a binary (zero vs.\ greater) and
a truncated (positive) part \citep[see][for an overview]{Winkelmann:2006}.
\begin{itemize}
\item Censored distribution \citep{Winkelmann:2006}: 
%In general, censoring data usually leads 
%to a point mass at the censoring point which has to be considered in 
%the modeling process. Therefore, censored distributions are applied as 
%they are able to capture this point mass. 
%In precipitation data, observations can not be negative, thus, left-censoring
%it at zero does not change the data. However, there is a naturally given
%point mass at zero (days without rain) making a censored distribution an 
%appropriate choice to model precipitation data.
For a chosen distribution with probability density function $f$ and
distribution function $F$ the log-likelihood function of the corresponding 
censored distribution is
%
\begin{equation}
\ell_{\mathrm{cens}}(\mu, \sigma; Y) = 
\begin{cases}
\log\left\{f(Y; \mu, \sigma) \right\}, & \text{if } Y > 0 \\[0.2cm]
\log\left\{F(0; \mu, \sigma) \right\}, & \text{if } Y = 0
\end{cases}
\end{equation}
%
as seen in Equation~2.1.1. of the paper for a Gaussian distribution.
For positive values of $Y$ the censored version of the density function
equals the original non-censored version. For $Y=0$ the censored density
function takes the values of the probability mass lying left of the 
censoring point. Therefore, the parameters $\mu$ and $\sigma$ control
both parts, the point mass and the positive part of the distribution.

\item Hurdle model / Two-part model (\cite{Cragg:1971}, \cite{Mullahy:1986}): 
An additional parameter $\nu$ describing the probability that there is 
any precipitation at all, i.e. $\nu = \mathbb{P}(Y>0)$, is included in the model.
The three-parametric log-likelihood function for the full hurdle model 
with parameter vector $\bm{\theta} = (\mu, \sigma, \nu)$ is
\begin{equation} \label{eq:hurdle}
\ell_{\mathrm{hurdle}}(\mu, \sigma, \nu; Y) = 
\begin{cases}
\log\left\{\nu \cdot \frac{f(Y; \mu, \sigma)}{1 - F(0; \mu, \sigma)} \right\}, & \text{if } Y > 0 \\[0.2cm]
\log(1 - \nu), & \text{if } Y = 0.
\end{cases}
\end{equation}
For $Y=0$ the probability that it won't rain is expressed by $1-\nu$. 
For $Y>0$ the probability that it will rain $\nu$ is multiplied with the 
left-truncated at zero density function. Contrary to the censored model, 
here it is only the parameter $\nu$ controlling the point mass while $\mu$ 
and $\sigma$ only influence the distribution of the positive observations.

>>>>

\title{Supplement 1: Different Response Distributions}

@@

\Abstract{
  In the case study presented in the main manuscript ``Distributional Regression
  Forests for Probabilistic Precipitation Forecasting in Complex Terrain''
  the novel distributional regression forests are based on a Gaussian response distribution,
  left-censored at zero to accomodate both the non-negativity of precipitation and
  the point mass at zero. To assess the goodness of fit of this response distribution
  this supplement employs the same evaluations as in the main manuscript but based on
  two other distributional assumptions: A logistic distribution, left-censored at zero,
  is employed to potentially better capture heavy tails -- and a two-part hurdle model
  combining a binary model for zero vs.\ positive precipitation and a Gaussian model,
  truncated at zero, for the positive precipitation observations. The latter could
  potentially better discriminate determinants of the occurrence of precipitation
  vs.\ the amount of precipitation given that it occurs.
}

@@

\section{Introduction}

This supplement extends the case study presented in Section~3.4 of the main
manuscript by carrying out the same evaluation at 15~stations using two
alternative response distributions for the power-transformed daily precipitation
sums.

To assess how appropriate the distributional assumption (left-censored Gaussian distribution
with censoring point zerol; or zero-censored for short) is, two questions are considered:
(1)~Are the tails of the distribution captured appropriately by a Gaussian distribution or
are \emph{heavier tails} required?
(2)~Is the point mass at zero (= no rain) driven by the same determinants as the
positive observations (= rain) or should this be captured by a \emph{separate parameter}?
For (1) we compare the zero-censored Gaussian with a \emph{zero-censored logistic} distribution and
for (2) a Gaussian \emph{hurdle model} (or two-part model with binary and zero-truncated parts)
is employed instead of the zero-censored Gaussian.

Just as in the main manuscript the performance of the distributional forest is also
compared with other commonly-used distributional regression models using the same
data set. Table~2 of the manuscript provides
an overview of all models considered and their specifications. As a reminder
the approaches are briefly summarized below.
%
\begin{itemize}

@@

\item \emph{EMOS} \citep{Gneiting+Raftery+Westveld:2005}: 
Ensemble model output statistics uses the ensemble 
mean of total precipitation as regressor in the location submodel 
and the corresponding ensemble standard deviation in the scale submodel.
The parameters are estimated by maximum likelihood.

\item \emph{Prespecified GAMLSS:} Smooth additive splines are selected for
the most relevant predictors based on meteorological expert knowledge
following \cite{Stauffer+Umlauf+Messner:2017}. The model is estimated by maximum
penalized likelihood using a backfitting algorithm \citep{Stasinopoulos+Rigby:2007}.

\item \emph{Boosted GAMLSS:} Smooth additive splines are selected
automatically from all available variables, using non-cyclic boosting
for parameter estimation \citep{Hofner+Mayr+Schmid:2016, Messner+Mayr+Zeileis:2017}.
The optimal stopping point for the iterations is determined based on 
a (computationally intensive) out-of-bag bootstrap estimate of the log-likelihood.

@@

\section{Alternative response distributions}

The main manuscript employes a Gaussian distribution with probability
density function $f$ (PDF) and cumulative distribution function $F$ (CDF)
given by
%
\begin{align}
f_{\mathrm{Gaussian}}(Y; \mu, \sigma) &= 
\frac{1}{\sigma} \cdot \phi\left(\frac{Y-\mu}{\sigma}\right) \\
F_{\mathrm{Gaussian}}(Y; \mu, \sigma) &=
\Phi\left(\frac{Y-\mu}{\sigma}\right)
\end{align}
where $\phi$ and $\Phi$ are the PDF and the CDF of the standard normal
distribution $\mathcal{N}(0,1)$.

An alternative distribution with several similar properties but with
heavier tails is the logistic distribution. The PDF and CDF are given by
%
\begin{align}
f_{\mathrm{logistic}}(Y; \mu, \sigma) &= 
\frac{\exp{\left(-\frac{Y - \mu}{\sigma}\right)}}
{\sigma \cdot \left(1 + \exp{\left(-\frac{Y - \mu}{\sigma}\right)}\right)^2} \\
F_{\mathrm{logistic}}(Y; \mu, \sigma) &=
\frac{1}{1+ \exp{\left(-\frac{Y - \mu}{\sigma}\right)}}.
\end{align}
%
Both distributions have to be adapted in our case study to accomodate that
precipitation is non-negative with a point mass at zero (= no rain).
Two common strategies for this are to employ either \emph{left-censoring at zero}
or a \emph{hurdle} (or two-part) approach, combining a binary (zero vs.\ greater) and
a truncated (positive) part \citep[see][for an overview of both techniques]{Winkelmann:2006}.
\begin{itemize}
\item \emph{Censored distribution:}
For a chosen distribution with PDF $f$ and CDF $F$ the log-likelihood function of the corresponding 
zero-censored distribution is
%
\begin{equation}
\ell_{\mathrm{cens}}(\mu, \sigma; Y) = 
\begin{cases}
\log\left\{f(Y; \mu, \sigma) \right\}, & \text{if } Y > 0 \\[0.2cm]
\log\left\{F(0; \mu, \sigma) \right\}, & \text{if } Y = 0
\end{cases}
\end{equation}
%
as stated in Equation~2.1.1 of the main manuscript for the Gaussian special case.
For positive values of $Y$ the censored version of the PDF
equals the original non-censored version. For $Y=0$ the censored density
function takes the values of the probability mass lying left of the 
censoring point. Therefore, the parameters $\mu$ and $\sigma$ control
both aspects, the point mass and the positive part of the distribution.

\item \emph{Hurdle model \citep[or two-part model,][]{Cragg:1971,Mullahy:1986}:}
An additional parameter $\nu$ describing the probability that there is 
any precipitation at all, i.e. $\nu = \mathbb{P}(Y>0)$, is included in the model.
The three-parameter log-likelihood function for the full hurdle model 
with parameter vector $\bm{\theta} = (\mu, \sigma, \nu)$ is
\begin{equation} \label{eq:hurdle}
\ell_{\mathrm{hurdle}}(\mu, \sigma, \nu; Y) = 
\begin{cases}
\log\left\{\nu \cdot \frac{f(Y; \mu, \sigma)}{1 - F(0; \mu, \sigma)} \right\}, & \text{if } Y > 0 \\[0.2cm]
\log(1 - \nu), & \text{if } Y = 0.
\end{cases}
\end{equation}
For $Y=0$ the probability for no rain is expressed by $1-\nu$. 
For $Y>0$ the probability that it will rain $\nu$ is multiplied with the 
left-truncated at zero density function. In contrast to the censored model, 
there is thus a dedicated parameter $\nu$ controlling the point mass while 
the other parameters $\mu$ and $\sigma$ only influence the distribution of
the positive observations.
